Effects of epidemiological structure on the transient evolution of HIV virulence

### 2016-06-24 ms/plos_latex.tex
more paper ...

In the particular example of HIV virulence eco-evolutionary dynamics
and epidemiological, we reach the slightly ironic conclusion that the
effort put into building a more realistic model essentially cancels
out, leaving us in the same position as if we had ignored the problem
of epidemiological complexity entirely and used a naive random-mixing
contact model. Herbeck \etal\ \cite{herbeck_hiv_2014} build a
network model of partnerships, but they set the ``partnership''
duration to 1 day --- very unrealistic in epidemic terms, but perhaps
actually more true to real-world HIV epidemiological dynamics than a
model with realistic partnership durations that neglects extra-pair
contact \cite{herbeck2016evolution}. Adding still further realism
might push the results farther away from the random-mixing model. For
example, our model forms partnerships randomly, and assumes that
extra-pair contact is randomly mixing across the population;
one could instead implement extra-pair contact as arising from
multiple concurrent partnerships (some, such as contact with sex
workers, of very short duration), or more structured partnership
formation (by age, ethnicity, or behaviour group). The effects of
other realistic complications such as explicit modeling of two
sexes (both in contact structure and differential transmission
probabilities), temporal and spatial variation in epidemic processes,
or presence of genetic variation in hosts are harder to predict.

Parameterization is one of the biggest challenges of epidemiological
modeling. In addition to following Champredon \etal\ \cite{champredon_hiv_2013} 
by doing Latin hypercube
sampling across a wide range of epidemiological parameters, we 
calibrated each set of parameters to the same initial epidemic
growth rate, chosen to match the results of previous models
\cite{shirreff_transmission_2011}.  The previous models drew their
parameters from previous cohort studies from the 1990s
\cite{wawer2005rates,hollingsworth_hiv1_2008}
rather than doing any explicit calibration to epidemic curves,
but they give reasonable order-of-magnitude
growth rates ($\approx 0.04~\textrm{year}^{-1}$)
for the early stages of the HIV epidemic (although considerably
lower than estimates of $\approx 0.07-0.1~\textrm{year}^{-1}$
based on population genetic reconstructions \cite{faria_early_2014}).
However, our reason for calibrating was not to match any
specific observed epidemic, but rather to make sure that
we were making meaningful comparisons across a range of
models with radically different epidemiological structures, and
hence involving different interpretations of the same quantitative
parameters.  For example, in models with instantaneous switching the
partnership dissolution rate $c$ is identical to the partnership
formation rate; in models with explicit partnership formation,
the partnership formation rate is also $c$ at equilibrium,
but might vary over the course of an epidemic.
It is not obvious whether models with equal parameters but
different structures should be directly compared; calibration
solves this problem.

More generally, any model that wants to be
taken seriously for management and forecasting purposes should
be calibrated to \emph{all} available data, using informative
priors to incorporate both realistic distributions of uncertainty
for all parameters from independent measurements \cite{elderd_uncertainty_2006}
and calibration from population-level observations of epidemic
trajectories. For practical (rather than exploratory) purposes,
such a procedure would also be an improvement on the common --- although not universal --- %
practice of assessing uncertainty over uniform ranges, rather than
distributions that allow more continuous variation in support over
the range of a parameter.

>>>>

In the particular example of HIV virulence eco-evolutionary dynamics
and epidemiological, we reach the slightly ironic conclusion that the
effort put into building a more realistic model essentially cancels
out, leaving us in the same position as if we had ignored the problem
of epidemiological complexity entirely and used a naive random-mixing
contact model. In Herbeck \etal's \cite{herbeck_hiv_2014} network model of partnerships, the partnership duration is set to 1 day --- very unrealistic in epidemic terms, but perhaps
actually more true to real-world HIV epidemiological dynamics than a
model with realistic partnership durations that neglects extra-pair
contact \cite{herbeck2016evolution}. Making the model even more realistic
might make the random-mixing model less appropriate. For
example, our model forms partnerships randomly, and assumes that
extra-pair contact is randomly mixing across the population;
one could instead model extra-pair contact as arising from
multiple concurrent partnerships (some, such as contact with sex
workers, of very short duration) and/or more structured partnership
formation (by age, ethnicity, or behaviour group). The effects of
other realistic complications such as explicit modeling of two
sexes (both in contact structure and differential transmission
probabilities), temporal and spatial variation in epidemic processes,
or presence of genetic variation in hosts are harder to predict.

Parameterization is one of the biggest challenges of epidemiological
modeling. In addition to following Champredon \etal\ \cite{champredon_hiv_2013} 
by doing Latin hypercube
sampling across a wide range of epidemiological parameters, we 
calibrated each set of parameters to the same initial epidemic
growth rate, chosen to match the results of previous models
\cite{shirreff_transmission_2011}.  Previous models 
in this area have drawn their
parameters from cohort studies from the 1990s
\cite{wawer2005rates,hollingsworth_hiv1_2008}
rather than doing any explicit calibration to epidemic curves,
but they give reasonable order-of-magnitude
growth rates ($\approx 0.04~\textrm{year}^{-1}$)
for the early stages of the HIV epidemic (although considerably
lower than estimates of $\approx 0.07-0.1~\textrm{year}^{-1}$
based on population genetic reconstructions \cite{faria_early_2014}).
However, our reason for calibrating was not to match any
specific observed epidemic, but rather to make sure that
we were making meaningful comparisons across a range of
models with radically different epidemiological structures, and
hence involving different interpretations of the same quantitative
parameters.  For example, in models with instantaneous switching the
partnership dissolution rate $c$ is identical to the partnership
formation rate; in models with explicit partnership formation,
the partnership formation rate is also $c$ at equilibrium,
but might vary over the course of an epidemic.
It is not obvious whether models with equal parameters but
different structures should be directly compared; calibration
solves this problem.

More generally, any model that wants to be
taken seriously for management and forecasting purposes should
be calibrated to \emph{all} available data, using informative
priors to incorporate both realistic distributions of uncertainty
for all parameters from independent measurements \cite{elderd_uncertainty_2006}
and calibration from population-level observations of epidemic
trajectories. Such a procedure would also be an improvement on the common --- although not universal --- %
practice, which we have followed here,
of assessing uncertainty over uniform ranges rather than
using distributions that allow more continuous variation in support over
the range of a parameter.
